Predator clusters

### 2023-08-18 reports/04_pred_clus.qmd
Cleaning up figures and tables

library(ggtree)
library(patchwork)
library(sf)
library(targets)
library(tidyterra)
library(tidyverse)
library(vegan)

tar_load_everything(store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
seaice_contours <- terra::unwrap(seaice_contours)
```

@@

indval_mtx <- sightings_indval$indval
indval_fmt <- apply(indval_mtx, 2, function(x) {
  ord <- order(x, decreasing = TRUE)
  result <- ifelse(x >= 0.25, 
         str_glue("{rownames(indval_mtx)} ({round(x * 100)}%)"),
         NA) %>% 
    `[`(ord) %>% 
    na.omit() %>% 
    paste(collapse = "<br>")
  ifelse(result == "", "None", result)
})
as_tibble(t(indval_fmt)) %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling()

@@

-   Cluster 1 = uncertain
-   Cluster 2 = pagophilic
-   Cluster 3 = open water

@@

How frequent are the different clusters and where were they found? The open water community was observed most often, followed by the marginal ice zone then the uncertain communities. The relative frequency of the open water community was greatest in 2012, when survey effort in the Bransfield Strait was most limited, and least in 2016. Generally, the open water community occupied the offshore regions, but in 2014 and 2015 it moved onto the shelf into the triangle between Elephant, Joinville, and George Islands.

@@

The uncertain community was most commonly observed in the south eastern part of the survey region, near Joinville Island. A particularly coherent group of sites in this community formed in 2013 between Joinville Island and the South Shetlands.

@@

ant_basemap(map_lim) +
  # Sea ice
  geom_tile(aes(x, y, fill = seaice_conc), 
            filter(seaice_conc_df, seaice_conc >= 0.65)) +
  scale_fill_distiller(lim = c(0.65, 1),
                       palette = "Blues",
                       na.value = "transparent") +
  geom_sf(data = seaice_contours, color = "darkblue") +
  # Predator clusters
  ggnewscale::new_scale_fill() +
  geom_sf(aes(fill = pred_clust), 
          stations_clust, 
          size = 2, 
          shape = 22) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(override.aes = list(size = 3))) +
  coord_ant(map_lim) +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

@@

```{r}
#| label: loading

ice_vector <- scores(nmds_envfit, 
                     display = "vectors",
                     choices = 1:3) %>% 
  as_tibble(rownames = "envvar") %>% 
  mutate(across(starts_with("NMDS"),
                ~ .x * ordiArrowMul(nmds_envfit))) %>% 
  filter(envvar == "ice_coverage")

@@

ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_point(data = year_centroids, 
             shape = 18,
             size = 2.5,
             color = "black",
             alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = Year),
                           year_centroids) +
  geom_segment(aes(xend = NMDS1, yend = NMDS2),
               ice_vector, 
               x = 0, y = 0,
               size = 1.5,
               color = "black",
               alpha = 0.65,
               arrow = arrow(angle = 20, length = unit(15, "points"))) +
  ggrepel::geom_text_repel(data = ice_vector,
                           label = "Ice coverage") +
  labs(color = "Predator cluster") +
  scale_color_brewer(palette = "Dark2") +
  theme_classic() +
  theme(legend.position = "bottom")

ggplot(nmds_df, aes(NMDS1, NMDS3)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_point(data = year_centroids, 
             shape = 18,
             size = 2.5,
             color = "black",
             alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = Year),
                           year_centroids) +
  geom_segment(aes(xend = NMDS1, yend = NMDS3),
               ice_vector, 
               x = 0, y = 0,
               size = 1.5,
               color = "black",
               alpha = 0.65,
               arrow = arrow(angle = 20, length = unit(15, "points"))) +
  ggrepel::geom_text_repel(data = ice_vector,
                           label = "Ice coverage") +
  labs(color = "Predator cluster") +
  scale_color_brewer(palette = "Dark2") +
  theme_classic() +
  theme(legend.position = "bottom")

ggplot(nmds_df, aes(NMDS2, NMDS3)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_point(data = year_centroids, 
             shape = 18,
             size = 2.5,
             color = "black",
             alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = Year),
                           year_centroids) +
  geom_segment(aes(xend = NMDS2, yend = NMDS3),
               ice_vector, 
               x = 0, y = 0,
               size = 1.5,
               color = "black",
               alpha = 0.65,
               arrow = arrow(angle = 20, length = unit(15, "points"))) +
  ggrepel::geom_text_repel(data = ice_vector,
                           label = "Ice coverage") +
  labs(color = "Predator cluster") +
  scale_color_brewer(palette = "Dark2") +
  theme_classic() +
  theme(legend.position = "bottom")
```

@@

Median ice coverage increased from the open water cluster (median 33.0%) to the marginal ice (59.6%) and uncertain (70.6%) clusters. This supports the labels of "open water" and "marginal ice" for clusters 3 and 2, respectively. Also provides evidence for the "pack ice" cluster.

@@

Predator/prey cluster associations were statistically significant overall ($\chi = 53.6, p<0.001$) as were 5 pairwise associations after applying Bonferroni correction. The marginal ice and uncertain predator communities were positively associated with macrozooplankton communities 3b (large krill, including *E. superba*) and 3a (an extremely diverse assemblage associated with cold, high salinity water), respectively. The open water predator community was positively associated with cluster 2b (*Thysanoessa macrura*) and negatively associated with 3a and 3b. Macrozooplankton cluster 1 (a ubiquitous cluster indicated by *Salpa* and *Clione*) was not significantly associated with any predator cluster. Neither was macrozooplankton cluster 2a (indicated by important prey species such as *E. frigida*, *E. triacantha*, myctophid larvae, and *Themisto gaudichaudii*), but it associated most often with the open water predator cluster.

@@

##

>>>>

library(ggtree)
library(gridExtra)
library(patchwork)
library(sf)
library(targets)
library(tidyterra)
library(tidyverse)
library(vegan)

tar_load_everything(store = here::here("_targets"))
source(here::here("R", "predators.R"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
seaice_contours <- terra::unwrap(seaice_contours)
```

@@

indval_mtx <- sightings_indval$indval
indval_fmt <- apply(indval_mtx, 2, function(x) {
  ord <- order(x, decreasing = TRUE)
  species <- code_to_common(rownames(indval_mtx))
  result <- ifelse(x >= 0.25, 
         str_glue("{species} ({round(x, 2)})"),
         NA) %>% 
    `[`(ord) %>% 
    na.omit() %>% 
    paste(collapse = "<br>")
  ifelse(result == "", "None", result)
})
as_tibble(t(indval_fmt)) %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling()

@@

-   Cluster 1 = pack ice
-   Cluster 2 = marginal ice
-   Cluster 3 = open water

@@

How frequent are the different clusters and where were they found? The open water community was observed most often, followed by the marginal ice zone then the pack ice communities. The relative frequency of the open water community was greatest in 2012, when survey effort in the Bransfield Strait was most limited, and least in 2016. Generally, the open water community occupied the offshore regions, but in 2014 and 2015 it moved onto the shelf into the triangle between Elephant, Joinville, and George Islands.

@@

The pack ice community was most commonly observed in the south eastern part of the survey region, near Joinville Island. A particularly coherent group of sites in this community formed in 2013 between Joinville Island and the South Shetlands.

@@

# Indicator species grob
indval_grob <- apply(sightings_indval$indval, 2, function(col) {
  # Extract indicator values >0.25 from each cluster's column and format with
  # species name
  col <- sort(col[col > 0.25], decreasing = TRUE)
  species <- code_to_common(names(col))
  if (length(col) == 0) {
    "None"
  } else {
    str_glue("{species} ({round(col, 2)})")
  }
}) %>% 
  # Create a 
  map(\(x) {
    nrow = max(map_dbl(., length))
    c(x, rep("", nrow - length(x)))
  }) %>% 
  as_tibble() %>% 
  set_names(levels(stations_clust$pred_clust)) %>% 
  tableGrob(
    rows = NULL,
    theme = ttheme_minimal(base_size = 6,
                           padding = unit(c(1, 1), "mm"),
                           core = list(fg_params = list(hjust = 0,
                                                        x = 0)),
                           colhead = list(fg_params = list(hjust = 0,
                                                           x = 0)))
  )

ant_basemap(map_lim) +
  # Sea ice
  geom_tile(aes(x, y, fill = seaice_conc), 
            filter(seaice_conc_df, between(seaice_conc, 0.65, 1)),
            alpha = 0.8) +
  scale_fill_distiller(lim = c(0.65, 1),
                       palette = "Blues",
                       na.value = "transparent") +
  # geom_sf(data = seaice_contours, color = "darkblue") +
  # Predator clusters
  ggnewscale::new_scale_fill() +
  geom_sf(aes(fill = pred_clust), 
          stations_clust, 
          size = 2, 
          shape = 22) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(override.aes = list(size = 3))) +
  coord_ant(map_lim) +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  inset_element(indval_grob, 
                0.67, 0, 1, 0.5, 
                align_to = "panel")
```

@@

```{r}
#| label: loading
#| fig-width: 9
#| fig-height: 5

@@

nmds_plot <- function(axis1, axis2) {
  ice_ordisurf <- ordisurf(
    nmds_sightings ~ nmds_env$ice_coverage,
    choices = c(axis1, axis2),
    plot = FALSE
  )

  x <- paste0("NMDS", axis1)
  y <- paste0("NMDS", axis2)

  ice_df <- expand_grid(
    axis1 = ice_ordisurf$grid$x,
    axis2 = ice_ordisurf$grid$y,
  ) %>% 
    cbind(as.numeric(ice_ordisurf$grid$z)) %>% 
    set_names(c(x, y, "ice_coverage")) %>% 
    mutate(ice_coverage = ice_coverage / 10) %>% 
    drop_na(ice_coverage)

  ggplot(nmds_df, aes(.data[[x]], .data[[y]])) +
    stat_contour(aes(z = .data$ice_coverage, 
                     color = after_stat(level)),
                 ice_df) +
    scale_color_distiller(
      "Ice coverage",
      palette = "Blues",
      labels = scales::percent,
      limits = c(0.25, 0.75),
      breaks = seq(0.25, 0.75, by = 0.1),
      guide = guide_colorbar(barwidth = unit(2, "in"),
                             direction = "horizontal")
    ) +
    ggnewscale::new_scale_color() +
    geom_point(aes(color = pred_clust), alpha = 0.8) +
    geom_point(data = year_centroids, 
               shape = 18,
               size = 2.5,
               color = "black",
               alpha = 0.85) +
    ggrepel::geom_text_repel(aes(label = Year),
                             year_centroids) +
    scale_color_brewer(
      palette = "Dark2",
      guide = guide_legend(override.aes = list(size = 2),
                           direction = "horizontal",
                           order = 1)
    ) +
    labs(color = "Predator cluster") +
    theme_classic()
}

map2(c(1, 1, 2), c(2, 3, 3), nmds_plot) %>% 
  reduce(`|`) +
  plot_annotation(tag_levels = 'A') +
  plot_layout(guides = "collect") &
    theme(legend.position = "bottom",
          legend.direction = "horizontal",
          plot.tag = element_text(face = "bold"))
```

@@

Median ice coverage increased from the open water cluster (median 33.0%) to the marginal ice (59.6%) and pack ice (70.6%) clusters. This supports the labels of "open water" and "marginal ice" for clusters 3 and 2, respectively. Also provides evidence for the "pack ice" cluster.

@@

Predator/prey cluster associations were statistically significant overall ($\chi = 53.6, p<0.001$) as were 5 pairwise associations after applying Bonferroni correction. The marginal ice and pack ice predator communities were positively associated with macrozooplankton communities 3b (large krill, including *E. superba*) and 3a (an extremely diverse assemblage associated with cold, high salinity water), respectively. The open water predator community was positively associated with cluster 2b (*Thysanoessa macrura*) and negatively associated with 3a and 3b. Macrozooplankton cluster 1 (a ubiquitous cluster indicated by *Salpa* and *Clione*) was not significantly associated with any predator cluster. Neither was macrozooplankton cluster 2a (indicated by important prey species such as *E. frigida*, *E. triacantha*, myctophid larvae, and *Themisto gaudichaudii*), but it associated most often with the open water predator cluster.

@@

```{r}
#| label: contingency

pred_prey_tbl <- with(stations_clust, 
                      table(pred_clust, `Winter Cluster factor`))
pred_prey_chisq$expected
pred_prey_chisq$observed

expand_grid(`Predator cluster` = levels(stations_clust$pred_clust),
            `Zooplankton cluster` = unique(stations_clust$`Winter Cluster factor`)) %>% 
  mutate(observed = map2_int(`Predator cluster`, 
                             `Zooplankton cluster`,
                             \(i, j) pred_prey_chisq$observed[i, j]),
         expected = map2_dbl(`Predator cluster`, 
                             `Zooplankton cluster`,
                             \(i, j) pred_prey_chisq$expected[i, j])) %>% 
  mutate(label = str_glue("{observed} ({round(expected, 1)})")) %>% 
  pivot_wider(id_cols = `Predator cluster`,
              names_from = `Zooplankton cluster`, 
              values_from = label) %>% 
  knitr::kable()
```
